Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-01-01 papers/wdmerger_I/paper.tex
Some more introduction cleanup

There are several observational reasons why double-degenerate systems are a promising progenitor system for at least a substantial fraction of normal SNe Ia. No conclusive evidence exists for a surviving companion star of a SN Ia; this is naturally explained by the DD model because both WDs are destroyed in the merger process. Similarly, pre-explosion images of the SN Ia systems have never clearly turned up a companion star, and in some cases a large fraction of the SD parameter space is excluded. Additionally, not enough progenitor systems are seen for the SD case to match the observed local SN Ia rate, whereas the number of white dwarf binaries may be sufficient to account for this rate. Finally, the DD model can naturally explain the fact that many SNe Ia are observed to occur at very long delay times after the stars were formed, since the progenitor systems only become active once both stars have evolved off the main sequence. A thorough review of the observational evidence about SNe Ia and further discussion of these ideas can be found in \cite{maoz:2014}.

The history of double degenerate theory can broadly be described as a series of three major paradigm shifts. The first attempts to model the results of the merger process came in the 1980s. \cite{nomotoiben:1985} demonstrated that if the secondary overflowed its Roche lobe and its mass accreted onto the primary near the Eddington rate, that off-center carbon ignition would occur. \cite{saionomoto:1985} tracked the evolution of the flame and found that it propagated quiescently into the center, converting the carbon-oxygen white dwarf into an oxygen-neon-magnesium white dwarf. This would then be followed by collapse into a neutron star -- a result with significantly different observational properties compared to a SN Ia. This scenario, termed accretion-induced collapse, would be avoided only if the accretion rate were well below the Eddington rate. \cite{tutukov_yungelson:1979} observed that this could happen if the mass loss from the secondary was higher than the Eddington rate and thus the accreted material formed an accretion disc, which might rain down on the primary more slowly. The main finding was that double degenerate systems would not obviously lead to Type Ia supernovae.

The first three-dimensional simulations of double degenerate systems were performed by \citet{benz:1990}, who used the smoothed particle hydrodynamics (SPH) method to simulate the merger process. They found that if the lower-mass secondary was close enough to the primary to begin mass transfer on a dynamical time scale, the secondary completely disrupted and formed a hot envelope on the primary, with a centrifugally-supported accretion disk surrounding the core and envelope. They noted that the envelope temperature was hot enough to generate carbon fusion but neglected reactions in their simulation. Later simulations of similar setups \citep{rasio_shapiro:1995,yoon:2007,loren-aguilar:2009,raskin:2012} corroborated this result. This validation of the prediction of \cite{tutukov_yungelson:1979} did not instill confidence in the community. \cite{mochkovitch_livio:1990} and \cite{livio:2000} observed that turbulent viscosity would be sufficiently large for angular momentum to be removed from the disk at a rate high enough to generate these troublesome accretion timescales. Based on all of this evidence, the review of \cite{hillebrandtniemeyer2000} argued that the model was only viable if the accretion-induced collapse problem could be avoided. Later work by \cite{shen:2012} and \cite{schwab:2012} used a more detailed treatment of the viscous transport in the outer regions of the remnant and found that while the centrifugally supported envelope would be converted into hot envelope material on a viscous timescale, their simulations still led to off-center carbon burning. \cite{vankerkwijk:2010} argued that equal-mass mergers would lead to the conditions necessary for carbon detonation in the center of the merged object, but \cite{shen:2012} questioned this for similar reasons related to how viscous transport would convert rotational motion into pressure support. \cite{zhu:2013} followed this with an expanded parameter space study and argued that many of their carbon-oxygen systems had the potential to detonate. The study of the long-term evolution of the remnants is thus still an open subject of research.

The most recent shift in perspective on this problem started with two series of papers appearing at roughly the same time. One began with \cite{pakmor:2010} and \cite{pakmor:2011}. This group used the SPH method to study the merger of equal-mass ($0.9\ M_\odot$) carbon-oxygen white dwarfs and found that a hotspot was generated near the surface of the primary white dwarf. They argued that this region had a temperature and density sufficient to trigger a thermonuclear detonation. They propagated this detonation throughout the system and found that it would observationally appear as a subluminous Type Ia supernova. This was the first time any simulation successfully reproduced at least some characteristics of a SN Ia. \cite{pakmor:2011} tried a few different mass combinations and found empirically that this would hold as long as the secondary was at least 80\% as massive as the primary. These events, where the merger process resulted in the detonation of the system during the merger coalescence -- avoiding the much longer time-scale evolution -- were termed ``violent'' mergers.

Around the same time, however, \cite{guillochon:2010} and \cite{dan:2011} came to significantly different conclusions. They pointed out that the previously mentioned simulations generally all shared a significant drawback, which was that their initial conditions were not carefully constructed. \cite{motl:2002}, \cite{dsouza:2006}, and \cite{motl:2007} (the first three-dimensional grid-based simulations of white dwarf mergers) pioneered the study of looking at the long-term dynamical evolution of merger scenarios after constructing exact equilibrium initial conditions. In contrast to the other studies, they found that paying close attention to the accuracy of the initial conditions of the simulation has important effects. In particular, earlier work placed the stars too close together and ignored the effects of tidal forces on changing the shape of the secondary, leading to the merger happening artificially too quickly. When the initial conditions are constructed in exact equilibrium, the system can be stable for tens of orbital periods, substantially changing the character of the mass transfer phase. However, one limitation of this work is that the authors used a polytropic equation of state and thus could not consider nuclear reactions. \cite{guillochon:2010} and \cite{dan:2011} improved on this using a realistic equation of state, a nuclear reaction network, and a similar approach to the equilibrium initial conditions, and found substantial agreement with the idea that mass transfer occurs in a stable manner over tens of orbital periods. They also found that, assuming the material accreted onto the surface of the primary was primarily helium, explosive surface detonations would occur as a result of accretion stream instabilities during the mass transfer phase prior to the full merger. This could trigger a double-detonation explosion and thus perhaps a SN Ia.

The most recent developments give some areas of agreement and some of remaining uncertainty. \cite{pakmor:2012} performed a merger scenario with a $1.1\ M_\odot$ and $0.9\ M_\odot$ setup, with better treatment of the initial conditions, and also found that the merger process happened over more than ten orbits. Nevertheless, they still found that a carbon-oxygen detonation would occur, in line with their earlier results. \cite{moll:2014} was also able to find a detonation in a similarly massive system. \cite{dan:2012} and \cite{dan:2014} performed a large sweep of the parameter space for merger pairs and found that pure carbon-oxygen systems would generally not lead to detonations and violent mergers except for the most massive systems. They did find that for systems containing helium, many of them would detonate and potentially lead to SNe Ia, either through the aforementioned instabilities in the accretion stream, or during the contact phase, similar to the violent carbon-oxygen mergers. \cite{pakmor:2013} added a thin helium shell on their primary white dwarf, and found that this robustly led to a detonation of the white dwarf. Thus at least a preliminary agreement may be that systems containing helium could robustly lead to events resembling SNe Ia, as well as very massive carbon-oxygen binaries.

>>>>

There are several observational reasons why double-degenerate systems are a promising progenitor system for at least a substantial fraction of normal SNe Ia. No conclusive evidence exists for a surviving companion star of a SN Ia; this is naturally explained by the DD model because both WDs are likely to be destroyed in the merger process. Similarly, pre-explosion images of the SN Ia systems have never clearly turned up a companion star, and in some cases a large fraction of the parameter space for the nature of the companion star is excluded. Additionally, not enough progenitor systems are seen for the SD case to match the observed local SN Ia rate, whereas the number of white dwarf binaries may be sufficient to account for this rate. Finally, the DD model can naturally explain the fact that many SNe Ia are observed to occur at very long delay times after the stars were formed, since the progenitor systems only become active once both stars have evolved off the main sequence. A thorough review of the observational evidence about SNe Ia and further discussion of these ideas can be found in \cite{maoz:2014}.

The history of double degenerate theory can broadly be described as a series of three major paradigm shifts. The first attempts to model the results of the merger process came in the 1980s. \cite{nomotoiben:1985} demonstrated that off-center carbon ignition would occur in the more massive white dwarf as it accreted mass near the Eddington rate from the less massive white dwarf overflowing its Roche lobe. \cite{saionomoto:1985} tracked the evolution of the flame and found that it propagated quiescently into the center, converting the carbon-oxygen white dwarf into an oxygen-neon-magnesium white dwarf. This would then be followed by collapse into a neutron star -- a result with significantly different observational properties compared to a SN Ia. This scenario, termed accretion-induced collapse, would be avoided only if the accretion rate were well below the Eddington rate. \cite{tutukov_yungelson:1979} observed that this could happen if the mass loss from the secondary was higher than the Eddington rate and thus the accreted material formed an accretion disc, which might rain down on the primary more slowly. The main finding was that double degenerate systems would not obviously lead to Type Ia supernovae.

The first three-dimensional simulations of double degenerate systems were performed by \citet{benz:1990}, who used the smoothed particle hydrodynamics (SPH) method to simulate the merger process. They found that if the lower-mass star (generally called the ``secondary'') was close enough to the more massive star (the ``primary'') to begin mass transfer on a dynamical time scale, the secondary completely disrupted and formed a hot envelope around the primary, with a centrifugally-supported accretion disk surrounding the core and envelope. They noted that the envelope temperature was hot enough to generate carbon fusion but neglected nuclear reactions in their simulation. Later simulations of similar setups \citep{rasio_shapiro:1995,yoon:2007,loren-aguilar:2009,raskin:2012} corroborated this result. This validation of the prediction of \cite{tutukov_yungelson:1979} did not instill confidence in the community. \cite{mochkovitch_livio:1990} and \cite{livio:2000} observed that turbulent viscosity would be sufficiently large for angular momentum to be removed from the disk at a rate high enough to generate the troublesome accretion timescales. Based on all of this evidence, the review of \cite{hillebrandtniemeyer2000} argued that the model was only viable if the accretion-induced collapse problem could be avoided. Later work by \cite{shen:2012} and \cite{schwab:2012} used a more detailed treatment of the viscous transport in the outer regions of the remnant and found that while the centrifugally supported envelope would be converted into hot envelope material on a viscous timescale, their simulations still led to off-center carbon burning. \cite{vankerkwijk:2010} argued that equal-mass mergers would lead to the conditions necessary for carbon detonation in the center of the merged object, but \cite{shen:2012} questioned this for similar reasons related to how viscous transport would convert rotational motion into pressure support. \cite{zhu:2013} followed this with an expanded parameter space study and argued that many of their carbon-oxygen systems had the potential to detonate. The study of the long-term evolution of the remnants is thus still an open subject of research.

The most recent shift in perspective on this problem started with two series of papers appearing at roughly the same time. One began with \cite{pakmor:2010} and \cite{pakmor:2011}. This group used the SPH method to study the merger of equal-mass ($0.9\ M_\odot$) carbon-oxygen white dwarfs and found that a hotspot was generated near the surface of the primary white dwarf. They argued that this region had a temperature and density sufficient to trigger a thermonuclear detonation. They propagated this detonation throughout the system and found that it would observationally appear as a subluminous Type Ia supernova. This was the first time a DD simulation successfully reproduced at least some characteristics of a SN Ia. \cite{pakmor:2011} tried a few different mass combinations and found empirically that this would hold as long as the secondary was at least 80\% as massive as the primary. These events, where the merger process resulted in the detonation of the system during the merger coalescence -- avoiding the much longer time-scale evolution -- were termed ``violent'' mergers.

Around the same time, however, \cite{guillochon:2010} and \cite{dan:2011} came to significantly different conclusions. They pointed out that the previously mentioned simulations generally all shared a significant drawback, which was that their initial conditions were not carefully constructed. \cite{motl:2002}, \cite{dsouza:2006}, and \cite{motl:2007} (the first three-dimensional grid-based simulations of white dwarf mergers) pioneered the study of looking at the long-term dynamical evolution of merger scenarios after constructing exact equilibrium initial conditions. In contrast to the other studies, they found that paying close attention to the accuracy of the initial conditions of the simulation has important effects. In particular, earlier work placed the stars too close together and ignored the effects of tidal forces on changing the shape of the secondary, leading to the merger happening artificially too quickly. When the initial conditions are constructed in exact equilibrium, the system can be stable for tens of orbital periods, substantially changing the character of the mass transfer phase. However, one limitation of this series of studies is that the authors used a polytropic equation of state and thus could not consider nuclear reactions. \cite{guillochon:2010} and \cite{dan:2011} improved on this using a realistic equation of state, a nuclear reaction network, and a similar approach to the equilibrium initial conditions, and found substantial agreement with the idea that mass transfer occurs in a stable manner over tens of orbital periods. They also found that, assuming the material accreted onto the surface of the primary was primarily helium, explosive surface detonations would occur as a result of accretion stream instabilities during the mass transfer phase prior to the full merger. This could trigger a double-detonation explosion and thus perhaps a SN Ia.

The very latest developments give some areas of agreement and some of remaining uncertainty. \cite{pakmor:2012} performed a merger scenario with a $1.1\ M_\odot$ and $0.9\ M_\odot$ setup, with better treatment of the initial conditions, and also found that the merger process happened over more than ten orbits. Nevertheless, they still found that a carbon-oxygen detonation would occur, in line with their earlier results. \cite{moll:2014} was also able to find a detonation in a similarly massive system. \cite{dan:2012} and \cite{dan:2014} performed a large sweep of the parameter space for merger pairs and found that pure carbon-oxygen systems would generally not lead to detonations and violent mergers except for the most massive systems. They did find that for systems containing helium, many of them would detonate and potentially lead to SNe Ia, either through the aforementioned instabilities in the accretion stream, or during the contact phase, similar to the violent carbon-oxygen mergers. \cite{pakmor:2013} added a thin helium shell on their primary white dwarf, and found that this robustly led to a detonation of the white dwarf. Thus at least a preliminary agreement may be that systems containing helium could robustly lead to events resembling SNe Ia, as well as very massive carbon-oxygen binaries.
